The Key to a Performant Data Parallel Compiler

### 2016-11-16 2017-spring-pldi/hsu-2017-key_data_parallel_compiler.tex
More clean-up and fixes

\title{The Key to a Performant Data Parallel Compiler}
\authorinfo{Aaron W. Hsu}{Indiana University, USA}{awhsu@indiana.edu}
\maketitle

\begin{abstract}
We present a language-driven strategy for the construction of compilers that are inherently data-parallel in their design and implementation. Using an encoding of the inter-node relationships between nodes in an AST called a Path Matrix, we demonstrate how an operator called the Key operator, that applies a function over groupings of array cells grouped and ordered by their keys, when used in conjunction with a Path Matrix, permits arbitrary computation over sub-trees of an AST using purely data-parallel array programming techniques. We discuss the use of these techniques when applied to an experimental commercial compiler called Co-dfns, a fully data-parallel compiler developed using the techniques discussed here. On transformation heavy operations such as expression flattening or function lifting, these techniques, when compared against pure and effectful recursive implementations, outperforms the CPU-based algorithms by a factor of 10 - 58 on the GPU. 
\end{abstract}

\category{D.3.4}{Programming Languages}{Processors}[Compilers]
\terms Algorithms, Performance, Design, Languages
\keywords APL, Parallel, Compilers, Key, Node Coordinates

@@

Compilers represent a peculiar class of tree/graph algorithms that greatly alter the underlying structure of the input graph into something very different in structure, but equivalent by some semantic interpretation. These algorithms are widely applicable, but have stubbornly resisted a general approach to parallelization. Most compilers are single-threaded or make very limited use of parallelism in an ad hoc way. The design and analysis of compilers almost universally deals with compilers as recursive traversals over ASTs. Such formulations usually include heavy reliance on branching, recursion, and sometimes very sophisticated control flow. All of this contributes to make it difficult to effectively parallelize such algo-rithms onto architectures that are sensitive to branching and recursion, such as GPGPUs or highly vectorised CPUs.

There has been some success in efficiently executing parts of a compiler on the GPU, such as the parser \cite{bunda1984apl}, tokenization \cite{bernecky2003tokenizer}, and certain compiler analyses \cite{prabhu2011eigencfa,mendez2012inclusion}. However, a generalized framework or strategy for parallel compilation remains elusive. One of the major difficulties is the recursive and highly branch dependent nature of the algorithms. These present challenges that must be overcome in vector-centric architectures.

We present a language-driven strategy for creating an inherently parallel compiler. We select a set of well known data-parallel primitives, and then restrict program construction to the composition of these operations into functions, without any forms of non-linear control flow, such as branching, recursion, or pattern matching. Programs constructed in such a style are data-parallel by construction. A key problem to the construction of compilers in such a restricted environment is dealing with the inherently recursive and nested structure of the AST. By working over a linearized representation of the AST as a matrix, combined with a structure we call a Path Matrix, and using data-parallel primitives, we are able to tame the recursive and nested nature of the AST into something that can be processed handily using only data-parallel, data-flow programs, without requiring branching or other forms of non-data parallel control flow. The methods are general and independent of the language being compiled.

@@

\noindent That is, we indent the expression by 6 spaces, followed by the value of the expression without indent. All expressions are evaluated right to left and all function application is infix; that is, a function application may apply a function (such as \verb;+;) to one or two arguments, called monadic and dyadic application respectively, which must appear on the right and (optionally) on the left of the function name, as in the above example. The Appendix provides a listing of all of the primitives used in the Co-dfns compiler, a selection of which are used here.

@@

\noindent We combine these to form the respective AST matrices. We write \verb;A,B; to catenate arrays \verb;A; and \verb;B; along their last axes and \verb;⍪A; to turn a vector into a 1-column matrix. Thus, the two AST matrices are given by the following expressions:

@@

\noindent In short, we care about the relative position of each node in a tree relative to any other. We define a path vector as a vector whose length is the depth of the tree. Its elements are natural numbers.  The count of non-zero elements in the vector is equal to the depth of that node in the tree. All zero elements appear after the non-zero elements. That is, a path is zero-padded on the right. When ordered lexicographically, the nodes for each path appear in order according to a depth-first pre-order traversal of the tree. Each path uniquely identifies a single node. Every ancestor’s path is a prefix of any child’s path, ignoring zeros. From the above it follows that every node is lexicographically greater than its left sibling and differs from it by exactly one non-zero element, and that this element is the last non-zero element in the path.

@@

\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\toprule
Pass & Description & Core Traversal Pattern & Uses coordinates? \\
\midrule
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting. & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes. & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals. & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time. &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables. &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables. &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions. &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{tab:passes}
\end{table*}

@@

Each of the following passes operates over an AST that we represent as the catenation of the field vectors above including the Node Coordinates.

@@

\noindent The function \verb;↓; converts the coordinate matrices from matrices to vectors of vectors, so that the coordinates can be catenated to the other fields, leading to a 4-column matrix for the ASTs, whose fourth column is a vector of vectors.

@@

Function lifting takes an AST with nested functions (functions appearing inside of other functions) and lifts these functions out of their local scopes and puts them all at the top level. Normally, function lifting must consider both the core operation of lifting function as well as the lexical scop-ing of the functions and variables that appear inside of a function. As seen below, our treatment uses a slightly different method.

If we use the Key operator (\verb;⌸;) with the pair function (\verb;{⍺ ⍵};) using \verb;Fk; as the keys and \verb;Fast; as the AST, we get the following:

@@

Notice that we have now grouped all of the relevant parts of the tree according to which nodes would appear in their respective functions after lifting. Refer to the original lifting example in the introduction to verify this. Indeed, the second row in the above example shows the internal function complete and ready to name. Each element in the second column corresponds to the body of one of our lifted functions. In the case of the first function, the outer function, we have a spurious function node in the body. This is intentional. When we lift these functions, we will replace each spurious function node with a variable node referring to the function’s generated name.

Each of these grouped sub-trees will be given a new local root \verb;F; node that is at the appropriate depth (0) and uses the key as its path vector (\verb;⊣;).

@@

\noindent At this point, the internal function nodes refer by path vector to the newly added local roots. This was down without traversing or maintaining these connections, but falls out as a natural artifact of using the Key (\verb;⌸;) operator. At this point we are ready to fix up the sub-tree to be correct. Firstly, we need a function to encapsulate the function header information.

@@

\noindent Next, we write a function to adjust the depths, which is one more than the difference between each node's depth and the depth of the first node of the sub-tree.

@@

\noindent Again, we can see immediately that we have grouped each set of nodes according to the expressions that are to be lifted. Just as in the case of function lifting, we can adjust the depths of each expression to the correct depth and we can replace each expression node with a variable reference based on that node’s coordinate. Each expression can be given a unique name based on its coordinate. A later compiler pass can reduce these names down to the minimum actually required to represent the expression if desired. As with the function lifting case, we need some functions to help abstract each new field. In our case, we use the same functions as with function lifting but alter the \verb;hd; and \verb;rf; functions to use the right node types:

@@

\noindent At this point, the expression lifting operation is then almost identical to that of function lifting. Notice that we reverse the 3-D matrix result at the end along the first access to ensure that order of evaluation is preserved correctly.

@@

@@

The Co-dfns compiler is an experimental commercial compiler funded by Dyalog, Ltd. \cite{hsu2014co,hsu2015accelerating} It is intended to provide increased scaling and performance for APL programmers using the dfns syntax (lexically scoped, functionally oriented APL). These techniques form one of the foundational elements to the architecture of the compiler, which is built in the style of a Nanopass \cite{keep2013nanopass} compiler, with very small passes chained together through composition, with the added requirements that each compiler pass must be fully data parallel.

@@

Table \ref{tab:passes} contains a selection of passes from the Co-dfns compiler and breaks down their features based on how they traverse the tree and whether or not they use the Key operator or Node Coordinates. The traversal pattern indicates the primary operator used to traverse the AST and perform the transformation, most of which are primitives; the Amend pattern is a ``replacement'' pattern where specific nodes are replaced in the AST with other nodes, and can be thought of as a slightly modified filter and map. The table orders the passes with front-end passes higher in the table and passes further down the compiler chain at the bottom of the table. These passes give a good idea of how the Key operator plays a part in the complete compiler as well as how Node Coordinates help to deal with complex sub-tree selections.

We make a few notes on the table and the analysis. The Key operator can be used with or without the node coordinates structure. In cases where the AST is sufficiently flat and arranged in a sufficiently convenient order, the grouping operations are very simple. In these cases, the Key operator amounts to a map operation over some selected sub-trees selected at a specific depth. These occur later in the compiler where the reader will note appearances of the Key operator without the use of Node Coordinates.

Likewise, node coordinates find their place early on in the compiler where it is necessary to deal with the more highly nested AST. They are useful for identifying this information regardless of whether the Key operator is used to handle the grouping or not. Thus passes like ``Drop Unreachable Code'' that remove whole sub-trees from the AST may use the Node Coordinates to do the grouping once the correct parent node has been identified.

In short, the goal of the structure and design of the compiler passes given in Table \ref{tab:passes} is to utilize Node Coordinates to remove as quickly as possible the nested structure of the AST into a flatter form that does not require complex analysis to traverse.

@@

We implemented the same algorithm using two traditional recursive methods. The first is a ``Functional'' analogue, that avoids mutation of the result and instead uses a pure-functional interface to pass nodes to be lifted back up the call stack. This is a very common method, taught in compiler courses that teach flattening passes, as well as used in various compilers. The benefit of the functional algorithm is the concision with which it may be written in most higher-level languages, particularly those which eschew mutatoin and favor a pure coding style. The disadvantage is that it has a higher memory usage requirement and may copy data as it moves up through the call stack to reassemble various containers for child nodes.

@@

Normally, one would test GPU oriented benchmarks based on their input sizes. However, in this case, the traditional algorithms and the GPU algorithms are written quite differently. Thus, we introduce two dimensions of input size in order to examine differences with more precision. Particularly, we benchmarked the GPU algorithm against the traditional methods by depth of the AST and Branching Factor. We chose to examine depth as a specific factor because the path matrix is inherently dependent on the depth. Furthermore, the mutation algorithm is optimized in such a way that fewer expression nodes may result in fewer writes to the accumulator, while our algorithm has no such optimization, and traverses the entire tree independent of the number of expression nodes.

We ran the benchmark to compute the time to flatten an AST on all combinations of depth and branching factor along the interval $[0,32)$ that were of a suitable size (that is, fit within the machine constraints). The GPU was an NVIDIA Tesla K40c, and the CPU was an Intel Xeon  E5-1620v3.

@@

There are many options for representing graphs on data-parallel architecture. Most of them tend to optimize traversal instead of transformation. Because of the transformation heavy nature of compiler-style algorithms, these other representations can introduce work that complicates both the algorithms themselves, as well as the work necessary to maintain those structures over the lifetime of the compilation phase.

@@

To illustrate the difference, consider a few other possible representations. One possibility is the use of a single parent vector for the entire AST. We could also work directly off of the depth vector or the traditional method of listing pointers to the children and/or parents. In the case of the expression flattening above, we could likely achieve equal or better performance from using a single parent vector for this one computation. However, would be increasing costs elsewhere, particularly in maintenance. A path matrix provides both parent information, as well as information about the depth of the node, the siblings, position in the tree, and a unique identifier in a single structure that is cache/memory friendly. Since many operations on trees are memory bound, this memory optimization becomes important. In the Co-dfns compiler, we are able to generate the path matrix once, at the beginning of the compilation, and pass them around freely, and even duplicate them when appropriate, to communicate information without requiring cross thread or non-local memory access patterns.

@@

We do not have a large scale understanding of the performance characteristics of these techniques in real life outside of daily use of the compiler. Further analysis is required to understand what sort of compiler optimizations and analyses are necessary to compile the Co-dfns compiler efficiently for the GPU and CPU. We currently lack a good comparison against existing techniques both for the compiler design and the generated code. Some existing compilers are able to compile variants of APL to target the GPU and the CPU, but there are subtle differences that require care when comparing generated code, and the architecture of these compilers is often quite different. Comparing the relative performance both of the compiler designs (human factors and execution speed) and generated code is important to understanding these techniques more fully.

@@

We have derived a method of performing computation over sub-trees selected on the basis of inter-node properties through the use of the Key (\verb;⌸;) operator and node coordinates, which enable local computation of these inter-node properties. This method is both general and direct, and when combined with traditional and more mundane array programming, suffices to implement the complete core of a compiler, modulo parsing and code generation. The method requires no special operations or unique special casing primitives in the language. Moreover, it is strictly data-parallel and data-flow, without any complex control flow, which results in exceptionally concise code. These methods permit a general, high-level approach to constructing a compiler that is parallel by construction by constricting the language to only a data parallel subset of a normal array language.

We have demonstrated the technique and the core insights behind the data structures involved. It presents a solution to a very old and traditional problem in a very uncommon light, by eschewing the common practices that underlie every other significant and general solution found in modern compilers today and replacing them with an entirely different paradigm centered on parallelism and aggregate operations. 
\balancecolumns

@@

\begin{table*}
\centering
\begin{tabular}{cllcll}
\toprule
Symbol                   & Monadic            & Dyadic  &
Symbol                   & Monadic            & Dyadic  \\
\midrule
\multicolumn{6}{c}{Scalar Functions} \\ 
\midrule
\texttt{+}               & Identity           & Plus (Add) &
\texttt{\textasciitilde} & Not                & \\
\texttt{-}               & Negative           & Minus (Subtract) &
\texttt{?}               & Roll               & \\
\texttt{×}               & Direction (Signum) & Times (Multiply) &
\texttt{∧}               &                    & And \\
\texttt{÷}               & Reciprocal         & Divide &
\texttt{∨}               &                    & Or \\
\texttt{|}               & Magnitude          & Residue (Modulo) &
\texttt{⍲}               &                    & Nand \\
\texttt{⌊}               & Floor              & Minimum &
\texttt{⍱}               &                    & Nor \\
\texttt{⌈}               & Ceiling            & Maximum &
\texttt{<}               &                    & Less \\
\texttt{*}               & Exponential        & Power &
\texttt{≤}               &                    & Less Or Equal \\
\texttt{⍟}               & Natural Logarithm  & Logarithm &
\texttt{=}               &                    & Equal \\
\texttt{○}               & Pi Times           & Circular (Trigonometric) &
\texttt{≥}               &                    & Greater Or Equal \\
\texttt{!}               & Factorial          & Binomial &
\texttt{>}               &                    & Greater \\
\texttt{≠}               &                    & Not Equal \\
\midrule
\multicolumn{3}{c}{Selection Mixed Functions} &
\multicolumn{3}{c}{Structural Mixed Functions} \\
\cmidrule(r){1-3} \cmidrule(l){4-6}
\texttt{⊃}               & Disclose & Pick &
\texttt{⍴} & Reshape       & \\
\texttt{↑}               &          & Take &
\texttt{,} & Ravel         & Catenate/Laminate \\
\texttt{↓}               &          & Drop &
\texttt{⍪} & Table         & Catenate First/Laminate \\
\texttt{/}               &          & Replicate &
\texttt{⌽} & Reverse       & Rotate \\
\texttt{⌿}               &          & Replicate First &
\texttt{⊖} & Reverse First & Rotate First \\
\texttt{\textbackslash}  &          & Expand &
\texttt{⍉} & Transpose     & Transpose \\
\texttt{⍀}               &          & Expand First &
\texttt{↑} & Mix           & \\
\texttt{\textasciitilde} &          & Without (Excluding) &
\texttt{↓} & Split         & \\
\texttt{∩}               &          & Intersection &
\texttt{⊂} & Enclose       & Partitioned Enclose \\
\texttt{∪}               & Unique   & Union &
\texttt{∊} & Enlist        & \\
\texttt{⊣}               & Same     & Left \\
\texttt{⊢}               & Identity & Right\\
\midrule
\multicolumn{3}{c}{Selector Mixed Functions} &
\multicolumn{3}{c}{Miscellaneous Mixed Functions} \\
\cmidrule(r){1-3} \cmidrule(l){4-6}
\texttt{⍳} & Index Generator & Index Of &
\texttt{⍴} & Shape         & \\
\texttt{∊} &                 & Membership &
\texttt{≡} & Depth         & Match \\
\texttt{⍋} & Grade Up        & Grade Up &
\texttt{≢} & Tally         & Not Match \\
\texttt{⍒} & Grade Down      & Grade Down &
\texttt{⍎} & Execute       & Execute \\
\texttt{?} &                 & Deal &
\texttt{⍕} & Format        & Format \\
\texttt{⍷} &                 & Find &
\texttt{⊥} &               & Decode (Base) \\
& & & \texttt{⊤} &               & Encode (Representation) \\
& & & \texttt{⌹} & Matrix Divide & Matrix Inverse \\
\end{tabular}
\caption{Primitive Functions}
\label{tab:scalarprims}
\end{table*}

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{⍨} & Commute & Swaps arguments or distributes
 right argument to both sides \\
\texttt{¨} & Each & Applies its operand point-wise over the left/right
 arguments \\
\texttt{/} & Reduce & Reduce along the last axis \\
\texttt{⌿} & Reduce First & Reduce along the first axis \\
\texttt{\textbackslash} & Scan & Scan along the last axis \\
\texttt{⍀} & Scan First & Scan along the first axis \\
\texttt{⌸} & Key & Apply operand once for each sub-array grouped by key \\
\end{tabular}
\caption{Primitive Monadic/Unary Operators, each takes a single left
operand and describes a function operating over one or two arguments}
\label{tab:adverbs}
\end{table*}

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{∘} & Compose & Composes two operands as in traditional mathematics \\
\texttt{.} & Inner Product & Inner product operation, e.g.
 \texttt{+.×} for matrix multiplication \\
\texttt{∘.} & Outer Product & Cartesian product or ``function table'' \\
\texttt{⍣} & Power & Iteration, Limited use only \\
\texttt{⍤} & Rank & Apply a function along cells of an array \\
\end{tabular}
\caption{Primitive Dyadic/Binary Operators, each takes a left
 and right operand and describes a function operating over one or two arguments}
\label{tab:conjunctions}
\end{table*}

>>>>

\title{The Key to a Performant Data Parallel Compiler}
%\authorinfo{Aaron W. Hsu}{Indiana University, USA}{awhsu@indiana.edu}
\authorinfo{}{}{}

\maketitle

\begin{abstract}
We present a language-driven strategy for the construction of compilers that are inherently data-parallel in their design and implementation. Using an encoding of the inter-node relationships between nodes in an AST called a Path Matrix, we demonstrate how an operator called the Key operator, that applies a function over groupings of array cells grouped and ordered by their keys, when used in conjunction with a Path Matrix, permits arbitrary computation over sub-trees of an AST using purely data-parallel array programming techniques. We discuss the use of these techniques when applied to an experimental commercial compiler called Co-dfns, a fully data-parallel compiler developed using the techniques discussed here. On transformation heavy operations such as expression flattening or function lifting, these techniques, when compared against pure and effectful recursive implementations, outperform the CPU-based algorithms by a factor of 10 - 58 on the GPU. 
\end{abstract}

\category{D.3.4}{Programming Languages}{Processors}[Compilers]
\terms Algorithms, Performance, Design, Languages
\keywords APL, Parallel, Compilers, Key, Paths

@@

Compilers represent a peculiar class of tree/graph algorithms that greatly alter the underlying structure of the input graph into something very different in structure, but equivalent by some semantic interpretation. These algorithms are widely applicable, but have stubbornly resisted a general approach to parallelization. Most compilers are single-threaded or make very limited use of parallelism in an ad hoc way. The design and analysis of compilers almost universally deals with compilers as recursive traversals over ASTs. Such formulations usually include heavy reliance on branching, recursion, and sometimes very sophisticated control flow. All of this contributes to make it difficult to effectively parallelize such algorithms onto architectures that are sensitive to branching and recursion, such as GPGPUs or highly vectorised CPUs.

There has been some success in efficiently executing parts of a compiler on the GPU, such as parsing \cite{bunda1984apl}, tokenization \cite{bernecky2003tokenizer}, and certain compiler analyses \cite{prabhu2011eigencfa,mendez2012inclusion}. However, a generalized framework or strategy for parallel compilation remains elusive. One of the major difficulties is the recursive and highly branch dependent nature of the algorithms. These present challenges that must be overcome in vector-centric architectures.

We present a language-driven strategy for creating an inherently parallel compiler. We select a set of well known data-parallel primitives, and then restrict program construction to the composition of these operations into functions, without any forms of non-linear control flow, such as branching, recursion, or pattern matching. Programs constructed in such a style are data-parallel by construction. A key problem to the construction of compilers in such a restricted environment is dealing with the inherently recursive and nested structure of the AST. By working over a linearized representation of the AST as a matrix, combined with a structure we call a Path Matrix, and using data-parallel primitives, we are able to tame the recursive and nested nature of the AST into something that can be processed handily using only data-parallel, data-flow programs, without requiring branching or other forms of non-data parallel control flow. The methods are general and independent of the transformation or computation.

@@

\noindent That is, we indent the expression by 6 spaces, followed by the value of the expression without indent. All expressions are evaluated right to left and all function application is infix; that is, a function application may apply a function (such as \verb;+;) to one or two arguments, called monadic and dyadic application respectively, which must appear on the right and (optionally) on the left of the function name, as in the above example. The appendix provides a listing of all of the primitives used in the Co-dfns compiler, a selection of which are used here.

@@

\noindent We combine these to form the respective AST matrices. We write \verb;A,B; to catenate arrays \verb;A; and \verb;B; along the last axis and \verb;⍪A; to turn a vector into a 1-column matrix. Thus, the two AST matrices are given by the following expressions:

@@

\noindent In short, we care about the relative position of each node in a tree relative to any other. We define a path vector as a vector whose length is the depth of the tree. Its elements are natural numbers.  The count of non-zero elements in the vector is equal to the depth of that node in the tree. All zero elements appear after the non-zero elements. That is, a path is zero-padded on the right. When ordered lexicographically, the nodes for each path appear in order according to a depth-first pre-order traversal of the tree. Each path uniquely identifies a single node. Every ancestor’s path is a prefix of any child’s path, ignoring zeros. From the above it follows that every node is lexicographically greater than its left sibling and differs only in the last non-zero element.

@@

@@

Each of the following passes operates over an AST represented as the catenation of the field vectors above including the Path Matrix.

@@

\noindent The function \verb;↓; converts the path matrices from matrices to vectors of vectors, so that the coordinates can be catenated to the other fields, leading to a 4-column matrix for the ASTs, whose fourth column is a vector of vectors.

@@

Function lifting takes an AST with nested functions (functions appearing inside of other functions) and lifts these functions out of their local scopes and puts them all at the top level. Normally, function lifting must consider both the core operation of lifting function as well as the lexical scoping of the functions and variables that appear inside of a function. As seen below, our treatment uses a slightly different method. By using the path matrix, we are able to maintain the relevant information throughout the compiler and avoid resolving scope until around the same time that we handle other variable resolution tasks.

If we use the Key operator (\verb;⌸;) with the pair function (\verb;{⍺ ⍵};) using \verb;Fk; as the keys and \verb;Fast; as the AST, we group all of the relevant parts of the tree according to which nodes would appear in their respective functions after lifting. Refer to the original AST given above to verify this. The second column in the following result shows the body of each function, complete and ready to name. Each element in the second column corresponds to the body of one of our to be lifted functions.

@@

In the case of the first function, the outer function, notice the spurious function node in the body. This is intentional. When we lift these functions, we will replace each spurious function node with a variable node referring to the function’s generated name.

Each of these grouped sub-trees will be given a new local root \verb;F; node that is at the appropriate depth (0) and uses the key as its path vector (\verb;⊣; or \verb;⍺;).

@@

\noindent At this point, the internal function nodes refer by path vector to the newly added local roots. This does not require traversing or maintaining these connections, but falls out as a natural artifact of using the Key (\verb;⌸;) operator. At this point we are ready to fix up the sub-tree to be correct. Firstly, we need a function to encapsulate the function header information.

@@

\noindent Next, we write a function to adjust the depths, which is one more than the difference between each node's depth and the depth of the first node of the sub-tree (\verb;n⌷⍤1; gives the $n$th column).

@@

\noindent Again, we can see immediately that we have grouped each set of nodes according to the expressions that are to be lifted. Just as in the case of function lifting, we can adjust the depths of each expression to the correct depth and we can replace each expression node with a reference based on that node’s path. Each expression has a unique name based on its path. A later compiler pass minimizes the names used in expressions (like register allocation). As with the function lifting case, we need some functions to help abstract each new field. In our case, we use the same functions as with function lifting but alter the \verb;hd; and \verb;rf; functions to use the right node types:

@@

\noindent The resulting expression lifting operation is almost identical to that of function lifting. Notice that we reverse the 3-D matrix result at the end along the first axis to preserve order of evaluation correctly.

@@

\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\toprule
Pass & Description & Core Traversal Pattern & Uses paths? \\
\midrule
Record Path Vectors & Adds a new field to each node containing that node's 
 path vector. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting. & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes. & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals. & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time. &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables. &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables. &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions. &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{tab:passes}
\end{table*}

@@

The Co-dfns compiler is an experimental commercial compiler \cite{hsu2014co,hsu2015accelerating}. It is intended to provide increased scaling and performance for APL programmers using the dfns syntax (lexically scoped, functionally oriented APL). These techniques form one of the foundational elements to the architecture of the compiler, which is built in the style of a Nanopass \cite{keep2013nanopass} compiler, with very small passes chained together through composition, with the added requirements that each compiler pass must be fully data parallel.

@@

Table \ref{tab:passes} contains a selection of passes from the Co-dfns compiler and breaks down their features based on how they traverse the tree and whether or not they use the Key operator or Path Matrix. The traversal pattern indicates the primary operator used to traverse the AST and perform the transformation, most of which are primitives; the Amend pattern is a ``replacement'' pattern where specific nodes are replaced in the AST with other nodes, and can be thought of as a slightly modified filter and map. The table orders the passes with front-end passes higher in the table and passes further down the compiler chain at the bottom of the table. These passes give a good idea of how the Key operator plays a part in the complete compiler as well as how Path vectors help to deal with complex sub-tree selections.

We make a few notes on the table and the analysis. The Key operator can be used with or without the path matrix. In cases where the AST is sufficiently flat and arranged in a sufficiently convenient order, the grouping operations are very simple. In these cases, the Key operator amounts to a map operation over some selected sub-trees selected at a specific depth. These occur later in the compiler where the reader will note appearances of the Key operator without the use of Path Vectors.

Likewise, path vectors find their place early on in the compiler where it is necessary to deal with the more highly nested AST. They are useful for identifying this information regardless of whether the Key operator is used to handle the grouping or not. Thus passes like ``Drop Unreachable Code'' that remove whole sub-trees from the AST may use the Path Matrix to do the grouping once the correct parent node has been identified.

In short, the goal of the structure and design of the compiler passes given in Table \ref{tab:passes} is to utilize Path Vectors to remove as quickly as possible the nested structure of the AST into a flatter form that does not require complex analysis to traverse.

@@

We implemented the same algorithm using two traditional recursive methods. The first is a ``Functional'' analogue, that avoids mutation of the result and instead uses a pure-functional interface to pass nodes to be lifted back up the call stack. This is a very common method, taught in compiler courses that teach flattening passes, as well as used in various compilers. The benefit of the functional algorithm is the concision with which it may be written in most higher-level languages, particularly those which eschew mutation and favor a pure coding style. The disadvantage is that it has a higher memory usage requirement and may copy data as it moves up through the call stack to reassemble various containers for child nodes.

@@

We use a slightly more complex, but optimized path matrix than the one given above. In particular, we require that each non-zero element of a path vector be a valid row index into the path matrix, and that this index be a reference to the corresponding ancestor of the given depth for the node corresponding to the given path vector. By ensuring that each element is a valid index, we are able to optimize the question of whether a given parent has a given node type. Instead of the more expensive  \verb;∧.(=∨0=⊢); based general solution, we can use the expression \verb;⌈/C×t=T[C]; to acquire suitable keys to group our sub-trees for a type \verb;t;, path matrix \verb;C;, and type vector \verb;T;. This optimized expression is not as general, but it covers a common case. The critical path for computing the nearest parent of a given type for the AST is less than depth of the AST, depending on how one implements the above expression.

We benchmarked the GPU algorithm against the traditional methods by depth of the AST and branching factor. We chose to examine depth as a specific factor because the path matrix is inherently dependent on the depth. Furthermore, the mutation algorithm is optimized in such a way that fewer expression nodes may result in fewer writes to the accumulator, while our algorithm has no such optimization, and traverses the entire tree independent of the number of expression nodes. We ran the benchmark to compute the time to flatten an AST on all combinations of depth and branching factor along the interval $[0,32)$ that were of a suitable size (that is, fit within the machine constraints). The GPU was an NVIDIA Tesla K40c, and the CPU was an Intel Xeon  E5-1620v3.

@@

There are many options for representing graphs on data-parallel architectures. Most of them tend to optimize traversal instead of transformation. Because of the transformation heavy nature of compiler-style algorithms, these other representations can introduce work that complicates both the algorithms themselves, as well as the work necessary to maintain those structures over the lifetime of the compilation phase.

@@

To illustrate the difference, consider a few other possible representations. One possibility is the use of a single parent vector for the entire AST. We could also work directly off of the depth vector or the traditional method of listing pointers to the children and/or parents. In the case of the expression flattening above, we could likely achieve equal or better performance from using a single parent vector for this one computation. However, we would increase costs elsewhere, particularly in maintenance. A path matrix provides both parent information, as well as information about the depth of the node, the siblings, position in the tree, and a unique identifier in a single structure that is cache/memory friendly. Since many operations on trees are memory bound, this memory optimization becomes important. In the Co-dfns compiler, we are able to generate the path matrix once, at the beginning of the compilation, and pass them around freely, and even duplicate them when appropriate, to communicate information without requiring cross thread or non-local memory access patterns.

@@

@@

We have derived a method of performing computation over sub-trees selected on the basis of inter-node properties through the use of the Key (\verb;⌸;) operator and path vectors, which enable local computation about these inter-node properties. This method is both general and direct, and when combined with traditional and more mundane array programming, suffices to implement the complete core of a compiler, modulo parsing and code generation. The method requires no special operations or unique special casing primitives in the language. Moreover, it is strictly data-parallel and data-flow, without any complex control flow, which results in exceptionally concise code. These methods permit a general, high-level approach to constructing a compiler that is parallel by construction by constricting the language to only a data parallel subset of a normal array language.

We have demonstrated the technique and the core insights behind the data structures involved. It presents a solution to a very old and traditional problem in a very uncommon light, by eschewing the common practices that underlie every other significant and general solution found in modern compilers today and replacing them with an entirely different paradigm centered on parallelism and aggregate operations. This new paradigm shows improved performance for transformation heavy algorithms which form the backbone of any non-trivial compiler on data-parallel or vector hardware such as GPUs. 
\balancecolumns

@@
